Accelerating Optimal Power Flow with GPUs: SIMD Abstraction of Nonlinear Programs and Condensed-Space Interior-Point Methods

### 2023-10-27 paper/main.tex
coin-or and AIChE talk

% \documentclass{IEEEtran4PSCC} % will be used for submission
% \author{
%   \IEEEauthorblockN{Sungho Shin and Mihai Anitescu}
%   \IEEEauthorblockA{
%     Mathematics and Computer Science Division\\
%     Argonne National Laboratory\\
%     Lemont, IL, USA\\
%     sshin@anl.gov, anitescu@mcs.anl.gov}
%   \and
%   \IEEEauthorblockN{François Pacaud}
%   \IEEEauthorblockA{Centre Automatique et Systèmes\\
%     Mines Paris - PSL \\
%     Paris, France\\
%     francois.pacaud@minesparis.psl.eu}
% }
% \psccfooter{%
%         \parbox{\textwidth}{\hrulefill \\ \small{23rd Power Systems Computation Conference} \hfill \begin{minipage}{0.2\textwidth}\centering \vspace*{4pt} \includegraphics[scale=0.06]{PSCC_logo.png}\\\small{PSCC 2024} \end{minipage} \hfill \small{Paris, France --- June 4 -- 7, 2024}}%
% }

\documentclass{IEEEtran} % for ArXiv submission
\author{
  Sungho Shin, François Pacaud, Mihai Anitescu
  \thanks{
    This material is based upon work supported by the U.S. Department of Energy, Office of Science, Office of Advanced Scientific Computing Research (ASCR) under Contract DE-AC02-06CH11347.
  }
  \thanks{
    Sungho Shin and Mihai Anitescu are with Mathematics and Computer Science Division,  Argonne National Laboratory, Lemont, IL, USA. e-mail: \url{sshin@anl.gov}, \url{anitescu@mcs.anl.gov}
  }
  \thanks{
    François Pacaud is with Centre Automatique et Systèmes, Mines Paris - PSL, Paris, France. e-mail: \url{francois.pacaud@minesparis.psl.eu}
  }
}

@@

% Set footer
\makeatletter
\let\old@ps@headings\ps@headings
\let\old@ps@IEEEtitlepagestyle\ps@IEEEtitlepagestyle
\def\psccfooter#1{%
    \def\ps@headings{%
        \old@ps@headings%
        \def\@oddfoot{\strut\hfill#1\hfill\strut}%
        \def\@evenfoot{\strut\hfill#1\hfill\strut}%
    }%
    \def\ps@IEEEtitlepagestyle{%
        \old@ps@IEEEtitlepagestyle%
        \def\@oddfoot{\strut\hfill#1\hfill\strut}%
        \def\@evenfoot{\strut\hfill#1\hfill\strut}%
    }%
    \ps@headings%
}
\makeatother

\usepackage{amsmath,amssymb,amsthm,amsfonts} % assumes amsmath package installed
\usepackage[linktocpage=true,colorlinks=true,linkcolor=blue,citecolor=blue,urlcolor=blue]{hyperref}
% \usepackage[letterpaper,margin=1in]{geometry}

\newtheorem{definition}{Definition}
\newtheorem{assumption}{Assumption}
\newtheorem{theorem}{Theorem}
\newtheorem{conjecture}{Conjecture}
\newtheorem{lemma}{Lemma}
\newtheorem{proposition}{Proposition}
\newtheorem{remark}{Remark}

\newcommand{\bx}{\boldsymbol{x}}
\newcommand{\be}{\boldsymbol{e}}
\newcommand{\blambda}{\boldsymbol{\lambda}}
\newcommand{\bLambda}{\boldsymbol{\Lambda}}
\newcommand{\bu}{\boldsymbol{u}}
\newcommand{\bw}{\boldsymbol{w}}
\newcommand{\by}{\boldsymbol{y}}
\newcommand{\bz}{\boldsymbol{z}}
\newcommand{\bV}{\boldsymbol{V}}
\newcommand{\bX}{\boldsymbol{X}}
\newcommand{\bY}{\boldsymbol{Y}}
\newcommand{\bZ}{\boldsymbol{Z}}
\newcommand{\bv}{\boldsymbol{v}}
\newcommand{\bxi}{\boldsymbol{\xi}}
\newcommand{\bpi}{\boldsymbol{\pi}}
\newcommand{\bphi}{\boldsymbol{\phi}}
\newcommand{\bbeta}{\boldsymbol{\eta}}
\newcommand{\bpsi}{\boldsymbol{\psi}}
\newcommand{\bzeta}{\boldsymbol{\zeta}}
\newcommand{\bmu}{\boldsymbol{\mu}}
\newcommand{\bq}{\boldsymbol{q}}
\newcommand{\bQ}{\boldsymbol{Q}}
\newcommand{\bK}{\boldsymbol{K}}
\newcommand{\bP}{\boldsymbol{P}}
\newcommand{\bS}{\boldsymbol{S}}
\newcommand{\bT}{\boldsymbol{T}}
\newcommand{\bF}{\boldsymbol{F}}
\newcommand{\bG}{\boldsymbol{G}}
\newcommand{\bd}{\boldsymbol{d}}
\newcommand{\bp}{\boldsymbol{p}}
\newcommand{\bff}{\boldsymbol{f}}
\newcommand{\bc}{\boldsymbol{c}}
\newcommand{\bg}{\boldsymbol{g}}
\newcommand{\bh}{\boldsymbol{h}}
\newcommand{\bA}{\boldsymbol{A}}
\newcommand{\bL}{\boldsymbol{L}}
\newcommand{\ba}{\boldsymbol{a}}
\newcommand{\bb}{\boldsymbol{b}}
\newcommand{\bB}{\boldsymbol{B}}
\newcommand{\bC}{\boldsymbol{C}}
\newcommand{\bE}{\boldsymbol{E}}
\newcommand{\bH}{\boldsymbol{H}}
\newcommand{\bR}{\boldsymbol{R}}
\newcommand{\bn}{\boldsymbol{n}}
\newcommand{\bm}{\boldsymbol{m}}
\newcommand{\br}{\boldsymbol{r}}
\newcommand{\bl}{\boldsymbol{l}}
\newcommand{\bI}{\boldsymbol{I}}
\newcommand{\osigma}{\overline{\sigma}}
\newcommand{\usigma}{\underline{\sigma}}
\newcommand{\oosigma}{\overline{\osigma}}
\newcommand{\uusigma}{\underline{\usigma}}
\newcommand{\olambda}{\overline{\lambda}}
\newcommand{\ulambda}{\underline{\lambda}}
\newcommand{\oolambda}{\overline{\olambda}}
\newcommand{\uulambda}{\underline{\ulambda}}
\newcommand{\bzero}{\boldsymbol{0}}
\newcommand{\dist}{\text{\normalfont dist}}
\newcommand{\st}{\mathop{\text{\normalfont s.t.}}}
\newcommand{\diag}{\mathop{\text{\normalfont diag}}}
\newcommand{\amin}{\mathop{\text{\normalfont argmin}}}
\newcommand{\ReH}{\mathop{\text{\normalfont ReH}}}
\newcommand{\bbZ}{\mathbb{Z}}
\newcommand{\cG}{\mathcal{G}}
\newcommand{\cV}{\mathcal{V}}
\newcommand{\cW}{\mathcal{W}}
\newcommand{\cA}{\mathcal{A}}
\newcommand{\cB}{\mathcal{B}}
\newcommand{\cL}{\mathcal{L}}
\newcommand{\cE}{\mathcal{E}}
\newcommand{\cD}{\mathcal{D}}
\newcommand{\cP}{\mathcal{P}}
\newcommand{\cQ}{\mathcal{Q}}
\newcommand{\cK}{\mathcal{K}}
\newcommand{\cM}{\mathcal{M}}
\newcommand{\cN}{\mathcal{N}}
\newcommand{\cI}{\mathcal{I}}
\newcommand{\cJ}{\mathcal{J}}
\newcommand{\cT}{\mathcal{T}}
\newcommand{\interior}{\mathop{\text{\normalfont interior}}}
\newcommand{\relint}{\mathop{\text{\normalfont relint}}}
\newcommand{\vertices}{\mathop{\text{\normalfont vertices}}}
\sloppy
\usepackage{graphicx}

\usepackage{enumitem}
\usepackage{mathtools}
\allowdisplaybreaks
\renewcommand{\theenumi}{(\alph{enumi})}
% \usepackage{parskip}

\usepackage{tikz}
\usepackage{lscape}
\usepackage{multirow}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{cuted}
\setlength\stripsep{3pt plus 1pt minus 1pt}

@@

\date{\small
  $^{1}$Mathematics and Computer Science Division, Argonne National Laboratory\\
  $^{2}$Centre Automatique et Syst\`{e}mes, Mines Paris - PSL
}
\begin{document}
\maketitle
\begin{abstract}
  This paper introduces a novel computational framework for solving
alternating current optimal power flow (ACOPF) problems using
graphics processing units (GPUs). While GPUs have demonstrated
remarkable performance in various computing domains, their application
in AC OPF has been limited due to challenges associated with porting
sparse automatic differentiation (AD) and sparse linear solver
routines to GPUs. We aim to address these issues with two key
strategies. First, we utilize a single-instruction, multiple-data
(SIMD) abstraction of nonlinear programs (NLP). This approach enables
the specification of model equations while preserving their
parallelizable structure, and in turn, facilitates the implementation
of AD routines that can exploit such structure. Second, we employ a
condensed-space interior-point method (IPM) with an inequality
relaxation strategy. This technique involves relaxing equality
constraints to inequalities and condensing the Karush-Kuhn-Tucker
system into a much smaller positive definite system. This strategy
offers the key advantage of being able to factorize the KKT matrix
without numerical pivoting, which in the past has hampered the parallelization of
the IPM algorithm. By combining these two strategies, we can perform the
majority of operations on GPUs while keeping the data residing in the
device memory only. Comprehensive numerical benchmark results showcase
the substantial computational advantage of our approach. Remarkably,
for solving large-scale AC OPF problems to a moderate accuracy, our
implementations—MadNLP.jl and ExaModels.jl—running on NVIDIA GPUs
achieve an order of magnitude speedup compared to state-of-the-art
tools running on contemporary CPUs.
\end{abstract}

\begin{IEEEkeywords}
  nonlinear programming, automatic differentiation, GPU computing,
optimal power flow
\end{IEEEkeywords}
\section{Introduction}

The adoption of GPUs in the mathematical programming community has remained
limited. Notably, nonlinear programming (NLP) remains dependent on
algorithms developed in the 1990s offering limited room for parallelism.
One of the primary challenges arises from the
automatic differentiation (AD) of sparse model equations and the
parallel factorization of indefinite sparse matrices, which are
commonly encountered within constrained numerical optimization tasks
\cite{anitescu2021targeting}. While GPU computation can trivially
accelerate several parts of the optimization process---especially
various internal computations within the optimization solver---the
sluggish data transfer between host and device memory hampers the
ad-hoc implementation of GPU accelerations (Fig. \ref{fig:memory}). To
fully leverage the potential offered by modern GPU hardware, it
becomes imperative to have a comprehensive computational framework for
optimization on GPUs. That is, we need an AD/algebraic
modeling framework, sparse linear solvers, and NLP solvers that can
operate entirely on the GPU. Specifically, for the best performance,
both the problem data and the solver's intermediate computational data
must be exclusively resident within the device memory, with the
majority of operations executed on the GPU.

\begin{figure}[t]
  \centering
  \begin{tikzpicture}[remember picture, scale=.9, font=\small]

    \fill[lightgray] (3.25,-.75) rectangle (4.75,4.75) node[black,midway,align=center,yshift=57.5] {PCIe\\Interface};
    \draw[gray,thick] (-.25,-.75) rectangle (2.75,4.75) node[black,midway,align=center,yshift=62] {Host (CPU)};
    \draw[gray,thick] (5.25,-.75) rectangle (8.25,4.75) node[black,midway,align=center,yshift=62] {Device (GPU)};

    \node[align=center] at (1.25, 3.75) {Data};
    \node[align=center] at (6.75, 3.75) {Data};

    % Host Memory
    \draw (0,-.5) rectangle (2.5,.5) node[midway] {Host Memory};

    % Device Memory
    \draw (5.5,-.5) rectangle (8,.5) node[midway] {Device Memory};

    % Arrow with Dashed Line
    \draw[<->, dashed] (2.5, 0) -- (5.5, 0) node[midway, above, align=center] {(slow)};
    \draw[<->, line width=1.5] (.25, .6) to [in=165, out =15] node[midway, above, align=center] {(fast)}(2.25, .6) ;
    \draw[<->, line width=1.5] (5.75, .6) to [in=165, out =15] node[midway, above, align=center] {(fast)}(7.75, .6) ;

    \def\rows{10}
    \def\cols{10}
    \def\elementwidth{2}
    \foreach \xshift/\yshift in {.25/1.5, 5.75/1.5} {

      % Draw Grid
      \foreach \i in {0,...,\rows} {
        \draw (\xshift, \yshift+\i*\elementwidth/\rows) -- (\xshift+\elementwidth, \yshift+\i*\elementwidth/\rows);
      }
      \foreach \i in {0,...,\cols} {
        \draw (\xshift+\i*\elementwidth/\cols, \yshift) -- (\xshift+\i*\elementwidth/\cols, \yshift+\elementwidth);
      }
    }
  \end{tikzpicture}
  \caption{A schematic description of host (CPU) and device (GPU) memory.}\label{fig:memory}
\end{figure}

This paper presents our approach to implementing a comprehensive
a computational framework for solving large-scale AC OPF problems on
NVIDIA GPUs, along with the associated software implementations:
ExaModels.jl \cite{simdiff}, an algebraic modeling/AD tool, and MadNLP.jl \cite{madnlp}, an
NLP solver. Our approach incorporates two novel strategies: (i) a
single-instruction, multiple-data (SIMD) abstraction of nonlinear
programs (NLPs), enabling streamlined parallel AD on GPUs, and (ii) a
condensed-space interior-point method (IPM) with an inequality
relaxation strategy, which facilitates the use of highly efficient
{\it refactorization} routines for sparse matrix factorization with
fixed pivot sequences.

While derivative evaluation can be generally cheaper than linear
algebra operations, our numerical results on AC OPF problems show that
AD often constitutes more than half of the total solver time when
using off-the-shelf AD implementations like
JuMP.jl~\cite{dunning2017jump} or
AMPL~\cite{fourer1990modeling}. Instead, our method leverages a
specialized AD implementation based on the SIMD abstraction of
NLPs. This abstraction allows to preserve the parallelizable structure
within the model equations, facilitating efficient derivative
evaluations on the GPU. The AC power flow model is particularly
well-suited for this abstraction as it involves repetitive expressions
for each component type (e.g., buses, lines, generators), and the
the number of computational patterns does not increase with the network's
size. Numerical results reported in this paper demonstrate that our
proposed strategies can achieve over 20 times speedup by running on
GPU. In comparison to general AD implementations on CPUs (such as AMPL
and JuMP.jl), our GPU-based differentiation method can be
approximately 500 times faster.

Linear algebra operations, especially sparse indefinite matrix
factorization, are typically the bottleneck in NLP solution methods.
Parallelizing this operation has been considered to be challenging,
primarily due to the need for numerical pivoting, which requires irregular
memory accesses and data movement \cite{swirydowicz2022linear}. However, when the matrix can be factorized without
numerical pivoting, a significant part of the operation can be
parallelized, and the numerical factorization can be efficiently
performed on GPUs. We develop a condensed-space IPM strategy that
allows the use of sparse matrix factorization routines without
numerical pivoting. This strategy relaxes equality constraints by
permitting small violations, which enables expressing the
Karush-Kuhn-Tucker (KKT) system entirely in the primal space through
the condensation procedure. Although this strategy is not new
\cite{nocedal2006numerical}, it has traditionally been considered less
efficient than the standard full-space method due to increased nonzero
entries in the KKT system. However, when implemented on GPUs, it
offers the key advantage of ensuring positive definiteness in the
condensed KKT system. This, in turn, allows for the utilization
of linear solvers with a fixed numerical pivot sequence (known as
refactorization), an efficient implementation of which is available in the CUDA library.
Although this method is susceptible to numerical stability issues due to
increased condition number in the KKT system, our results demonstrate
that the solver is robust enough to solve problems with a relative
accuracy of $\epsilon_{\text{mach}}^{1/4}\approx 10^{-4}$.

We present numerical benchmark results to showcase the efficiency of
our method, utilizing our two packages: MadNLP.jl and
ExaModels.jl. The solution of the KKT system is performed using the
external cuSOLVER library. To assess the performance of our method, we
compare it against standard CPU approaches using the data available in
pglib-opf \cite{babaeinejadsarookolaee2019power}.  Our benchmark results
demonstrate that our proposed computational framework has significant
potential for accelerating the solution of AC OPF problems, especially
when a moderate accuracy (e.g., $10^{-4}$) is sufficient.
Notably, when running on NVIDIA GPUs, our method achieves a
4x speedup compared to our solver running on CPU for
the largest instance. Moreover, for the same instance, our approach
surpasses the performance of existing tools (such as Ipopt interfaced
with JuMP.jl) by an order of magnitude. This finding underscores the
importance of harnessing the power of GPUs for tackling
the computational challenges in power systems.

\paragraph*{Contributions}
We present, for the first time, a sparse nonlinear optimization
solution framework that can run entirely on GPU, with all the
performance-critical data arrays residing exclusively on
GPUs. Additionally, we introduce the concept of SIMD abstraction for
NLP problems, which results in an efficient implementation of
GPU-accelerated parallel AD. Furthermore, we propose the condensed IPM
with an inequality relaxation strategy for the first time, enabling
the treatment of the KKT systems of NLPs without numerical pivoting,
thus allowing the solution of sparse, large-scale NLPs (with a
prominent example being AC OPFs) on GPUs.

>>>>

\documentclass{IEEEtran4PSCC} % will be used for submission
\author{
  \IEEEauthorblockN{Sungho Shin and Mihai Anitescu}
  \IEEEauthorblockA{
    Mathematics and Computer Science Division\\
    Argonne National Laboratory,
    Lemont, IL, USA\\
    sshin@anl.gov, anitescu@mcs.anl.gov}
  \and
  \IEEEauthorblockN{François Pacaud}
  \IEEEauthorblockA{Centre Automatique et Systèmes\\
    Mines Paris - PSL,
    Paris, France\\
    francois.pacaud@minesparis.psl.eu}
}
% Set footer
\makeatletter
\let\old@ps@headings\ps@headings
\let\old@ps@IEEEtitlepagestyle\ps@IEEEtitlepagestyle
\def\psccfooter#1{%
    \def\ps@headings{%
        \old@ps@headings%
        \def\@oddfoot{\strut\hfill#1\hfill\strut}%
        \def\@evenfoot{\strut\hfill#1\hfill\strut}%
    }%
    \def\ps@IEEEtitlepagestyle{%
        \old@ps@IEEEtitlepagestyle%
        \def\@oddfoot{\strut\hfill#1\hfill\strut}%
        \def\@evenfoot{\strut\hfill#1\hfill\strut}%
    }%
    \ps@headings%
}
\makeatother

\psccfooter{%
        \parbox{\textwidth}{\hrulefill \\ \small{23rd Power Systems Computation Conference} \hfill \begin{minipage}{0.2\textwidth}\centering \vspace*{4pt} \includegraphics[scale=0.06]{PSCC_logo.png}\\\small{PSCC 2024} \end{minipage} \hfill \small{Paris, France --- June 4 -- 7, 2024}}%
}

% \documentclass{IEEEtran} % for ArXiv submission
% \author{
%   Sungho Shin, François Pacaud, Mihai Anitescu
%   \thanks{
%     This material is based upon work supported by the U.S. Department of Energy, Office of Science, Office of Advanced Scientific Computing Research (ASCR) under Contract DE-AC02-06CH11347.
%   }
%   \thanks{
%     Sungho Shin and Mihai Anitescu are with Mathematics and Computer Science Division,  Argonne National Laboratory, Lemont, IL, USA. e-mail: \url{sshin@anl.gov}, \url{anitescu@mcs.anl.gov}
%   }
%   \thanks{
%     François Pacaud is with Centre Automatique et Systèmes, Mines Paris - PSL, Paris, France. e-mail: \url{francois.pacaud@minesparis.psl.eu}
%   }
% }

@@

% % Set footer
% \makeatletter
% \let\old@ps@headings\ps@headings
% \let\old@ps@IEEEtitlepagestyle\ps@IEEEtitlepagestyle
% \def\psccfooter#1{%
%     \def\ps@headings{%
%         \old@ps@headings%
%         \def\@oddfoot{\strut\hfill#1\hfill\strut}%
%         \def\@evenfoot{\strut\hfill#1\hfill\strut}%
%     }%
%     \def\ps@IEEEtitlepagestyle{%
%         \old@ps@IEEEtitlepagestyle%
%         \def\@oddfoot{\strut\hfill#1\hfill\strut}%
%         \def\@evenfoot{\strut\hfill#1\hfill\strut}%
%     }%
%     \ps@headings%
% }
% \makeatother

\usepackage{graphicx}
\usepackage{enumitem}
\usepackage{mathtools}
\usepackage{tikz}
\usepackage{lscape}
\usepackage{multirow}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{cuted}
\usepackage{amsmath,amssymb,amsthm,amsfonts} % assumes amsmath package installed
\usepackage[linktocpage=true,colorlinks=true,linkcolor=blue,citecolor=blue,urlcolor=blue]{hyperref}

% \newtheorem{definition}{Definition}
% \newtheorem{assumption}{Assumption}
% \newtheorem{theorem}{Theorem}
% \newtheorem{conjecture}{Conjecture}
% \newtheorem{lemma}{Lemma}
% \newtheorem{proposition}{Proposition}
% \newtheorem{remark}{Remark}

\newcommand{\bx}{\boldsymbol{x}}
\newcommand{\be}{\boldsymbol{e}}
\newcommand{\blambda}{\boldsymbol{\lambda}}
\newcommand{\bLambda}{\boldsymbol{\Lambda}}
\newcommand{\bu}{\boldsymbol{u}}
\newcommand{\bw}{\boldsymbol{w}}
\newcommand{\by}{\boldsymbol{y}}
\newcommand{\bz}{\boldsymbol{z}}
\newcommand{\bV}{\boldsymbol{V}}
\newcommand{\bX}{\boldsymbol{X}}
\newcommand{\bY}{\boldsymbol{Y}}
\newcommand{\bZ}{\boldsymbol{Z}}
\newcommand{\bv}{\boldsymbol{v}}
\newcommand{\bxi}{\boldsymbol{\xi}}
\newcommand{\bpi}{\boldsymbol{\pi}}
\newcommand{\bphi}{\boldsymbol{\phi}}
\newcommand{\bbeta}{\boldsymbol{\eta}}
\newcommand{\bpsi}{\boldsymbol{\psi}}
\newcommand{\bzeta}{\boldsymbol{\zeta}}
\newcommand{\bmu}{\boldsymbol{\mu}}
\newcommand{\bq}{\boldsymbol{q}}
\newcommand{\bQ}{\boldsymbol{Q}}
\newcommand{\bK}{\boldsymbol{K}}
\newcommand{\bP}{\boldsymbol{P}}
\newcommand{\bS}{\boldsymbol{S}}
\newcommand{\bT}{\boldsymbol{T}}
\newcommand{\bF}{\boldsymbol{F}}
\newcommand{\bG}{\boldsymbol{G}}
\newcommand{\bd}{\boldsymbol{d}}
\newcommand{\bp}{\boldsymbol{p}}
\newcommand{\bff}{\boldsymbol{f}}
\newcommand{\bc}{\boldsymbol{c}}
\newcommand{\bg}{\boldsymbol{g}}
\newcommand{\bh}{\boldsymbol{h}}
\newcommand{\bA}{\boldsymbol{A}}
\newcommand{\bL}{\boldsymbol{L}}
\newcommand{\ba}{\boldsymbol{a}}
\newcommand{\bb}{\boldsymbol{b}}
\newcommand{\bB}{\boldsymbol{B}}
\newcommand{\bC}{\boldsymbol{C}}
\newcommand{\bE}{\boldsymbol{E}}
\newcommand{\bH}{\boldsymbol{H}}
\newcommand{\bR}{\boldsymbol{R}}
\newcommand{\bn}{\boldsymbol{n}}
\newcommand{\bm}{\boldsymbol{m}}
\newcommand{\br}{\boldsymbol{r}}
\newcommand{\bl}{\boldsymbol{l}}
\newcommand{\bI}{\boldsymbol{I}}
\newcommand{\osigma}{\overline{\sigma}}
\newcommand{\usigma}{\underline{\sigma}}
\newcommand{\oosigma}{\overline{\osigma}}
\newcommand{\uusigma}{\underline{\usigma}}
\newcommand{\olambda}{\overline{\lambda}}
\newcommand{\ulambda}{\underline{\lambda}}
\newcommand{\oolambda}{\overline{\olambda}}
\newcommand{\uulambda}{\underline{\ulambda}}
\newcommand{\bzero}{\boldsymbol{0}}
\newcommand{\dist}{\text{\normalfont dist}}
\newcommand{\st}{\mathop{\text{\normalfont s.t.}}}
\newcommand{\diag}{\mathop{\text{\normalfont diag}}}
\newcommand{\amin}{\mathop{\text{\normalfont argmin}}}
\newcommand{\ReH}{\mathop{\text{\normalfont ReH}}}
\newcommand{\bbZ}{\mathbb{Z}}
\newcommand{\cG}{\mathcal{G}}
\newcommand{\cV}{\mathcal{V}}
\newcommand{\cW}{\mathcal{W}}
\newcommand{\cA}{\mathcal{A}}
\newcommand{\cB}{\mathcal{B}}
\newcommand{\cL}{\mathcal{L}}
\newcommand{\cE}{\mathcal{E}}
\newcommand{\cD}{\mathcal{D}}
\newcommand{\cP}{\mathcal{P}}
\newcommand{\cQ}{\mathcal{Q}}
\newcommand{\cK}{\mathcal{K}}
\newcommand{\cM}{\mathcal{M}}
\newcommand{\cN}{\mathcal{N}}
\newcommand{\cI}{\mathcal{I}}
\newcommand{\cJ}{\mathcal{J}}
\newcommand{\cT}{\mathcal{T}}
\newcommand{\interior}{\mathop{\text{\normalfont interior}}}
\newcommand{\relint}{\mathop{\text{\normalfont relint}}}
\newcommand{\vertices}{\mathop{\text{\normalfont vertices}}}
\sloppy
\allowdisplaybreaks
\renewcommand{\theenumi}{(\alph{enumi})}
% \usepackage{parskip}

\setlength\stripsep{3pt plus 1pt minus 1pt}

@@

\begin{document}
\maketitle

\begin{abstract}
  This paper introduces a framework for solving alternating current optimal power flow (ACOPF) problems using graphics processing units (GPUs). While GPUs have demonstrated remarkable performance in various computing domains, their application in ACOPF has been limited due to challenges associated with porting sparse automatic differentiation (AD) and sparse linear solver routines to GPUs. We  address these issues with two key strategies. First, we utilize a single-instruction, multiple-data  abstraction of nonlinear programs. This approach enables the specification of model equations while preserving their parallelizable structure and, in turn, facilitates the parallel AD implementation. Second, we employ a condensed-space interior-point method (IPM) with an inequality relaxation. This technique involves condensing the Karush--Kuhn--Tucker (KKT) system into a positive definite system. This strategy offers the key advantage of being able to factorize the KKT matrix without numerical pivoting, which has hampered the parallelization of the IPM algorithm. By combining these strategies, we can perform the majority of operations on GPUs while keeping the data residing in the device memory only. Comprehensive numerical benchmark results showcase the advantage of our approach. Remarkably, our implementations—MadNLP.jl and ExaModels.jl—running on NVIDIA GPUs achieve an order of magnitude speedup compared with state-of-the-art tools running on contemporary CPUs.
\end{abstract}

\begin{IEEEkeywords}
  nonlinear programming, automatic differentiation, GPU computing,
optimal power flow
\end{IEEEkeywords}

\section{Introduction}

The adoption of graphics processing units (GPUs) in the mathematical programming community has remained
limited. Notably, nonlinear programming (NLP) remains dependent on
algorithms developed in the 1990s offering limited room for parallelism.
One of the primary challenges arises from the
automatic differentiation (AD) of sparse model equations and the
parallel factorization of indefinite sparse matrices, which are
commonly encountered within constrained numerical optimization tasks
\cite{anitescu2021targeting}. While GPU computation can trivially
accelerate several parts of the optimization process---especially
various internal computations within the optimization solver---the
sluggish data transfer between host and device memory hampers the
ad hoc implementation of GPU accelerations. To
fully leverage the potential offered by modern GPU hardware, a comprehensive computational framework for
optimization on GPUs is imperative. That is, we need an AD/algebraic
modeling framework, sparse linear solvers, and NLP solvers that can
operate entirely on the GPU. Specifically, for the best performance,
both the problem data and the solver's intermediate computational data
must be exclusively resident within the device memory, with the
majority of operations executed on the GPU.

% \begin{figure}[t]
%   \centering
%   \begin{tikzpicture}[remember picture, scale=.9, font=\small]

%     \fill[lightgray] (3.25,-.75) rectangle (4.75,4.75) node[black,midway,align=center,yshift=57.5] {PCIe\\Interface};
%     \draw[gray,thick] (-.25,-.75) rectangle (2.75,4.75) node[black,midway,align=center,yshift=62] {Host (CPU)};
%     \draw[gray,thick] (5.25,-.75) rectangle (8.25,4.75) node[black,midway,align=center,yshift=62] {Device (GPU)};

%     \node[align=center] at (1.25, 3.75) {Data};
%     \node[align=center] at (6.75, 3.75) {Data};

%     % Host Memory
%     \draw (0,-.5) rectangle (2.5,.5) node[midway] {Host Memory};

%     % Device Memory
%     \draw (5.5,-.5) rectangle (8,.5) node[midway] {Device Memory};

%     % Arrow with Dashed Line
%     \draw[<->, dashed] (2.5, 0) -- (5.5, 0) node[midway, above, align=center] {(slow)};
%     \draw[<->, line width=1.5] (.25, .6) to [in=165, out =15] node[midway, above, align=center] {(fast)}(2.25, .6) ;
%     \draw[<->, line width=1.5] (5.75, .6) to [in=165, out =15] node[midway, above, align=center] {(fast)}(7.75, .6) ;

%     \def\rows{10}
%     \def\cols{10}
%     \def\elementwidth{2}
%     \foreach \xshift/\yshift in {.25/1.5, 5.75/1.5} {

%       % Draw Grid
%       \foreach \i in {0,...,\rows} {
%         \draw (\xshift, \yshift+\i*\elementwidth/\rows) -- (\xshift+\elementwidth, \yshift+\i*\elementwidth/\rows);
%       }
%       \foreach \i in {0,...,\cols} {
%         \draw (\xshift+\i*\elementwidth/\cols, \yshift) -- (\xshift+\i*\elementwidth/\cols, \yshift+\elementwidth);
%       }
%     }
%   \end{tikzpicture}
%   \caption{A schematic description of host (CPU) and device (GPU) memory.}\label{fig:memory}
%   \vspace{-.15in}
% \end{figure}

This paper presents our approach to implementing a comprehensive
computational framework for solving large-scale alternating current optimal power flow (ACOPF) problems on
NVIDIA GPUs, along with the associated software implementations:
ExaModels.jl \cite{simdiff}, an algebraic modeling/AD tool, and MadNLP.jl \cite{madnlp}, an
NLP solver. Our approach incorporates two novel strategies: (i) a
single-instruction, multiple-data (SIMD) abstraction of nonlinear
programs, enabling streamlined parallel AD on GPUs, and (ii) a
condensed-space interior-point method (IPM) with an inequality
relaxation strategy, which facilitates the use of highly efficient
{\it refactorization} routines for sparse matrix factorization with
fixed pivot sequences.

While derivative evaluation can be generally cheaper than linear
algebra operations, our numerical results on ACOPF problems show that
AD often constitutes more than half of the total solver time when
using off-the-shelf AD implementations such as
JuMP.jl~\cite{dunning2017jump} or
AMPL~\cite{fourer1990modeling}. Instead, our method leverages a
specialized AD implementation based on the SIMD abstraction of
NLPs. This abstraction allows us to preserve the parallelizable structure
within the model equations, facilitating efficient derivative
evaluations on the GPU. The AC power flow model is particularly
well suited for this abstraction because it involves repetitive expressions
for each component type (e.g., buses, lines, generators) and the
the number of computational patterns does not increase with the network's
size. Numerical results reported in this paper demonstrate that our
proposed strategies can achieve over 20 times speedup by running on
the GPU. Compared with general AD implementations on CPUs (such as AMPL
and JuMP.jl), our GPU-based differentiation method can be
approximately 500 times faster.

Linear algebra operations, especially sparse indefinite matrix
factorization, are typically the bottleneck in NLP solution methods.
Parallelizing this operation has been considered challenging,
primarily because of the need for numerical pivoting, which requires irregular
memory accesses and data movement \cite{swirydowicz2022linear}. However, when the matrix can be factorized without
numerical pivoting, a significant part of the operation can be
parallelized, and the numerical factorization can be efficiently
performed on GPUs. We develop a condensed-space IPM strategy that
allows the use of sparse matrix factorization routines without
numerical pivoting. This strategy relaxes equality constraints by
permitting small violations that enable expressing the
Karush--Kuhn--Tucker (KKT) system entirely in the primal space through
the condensation procedure. Although this strategy is not new
\cite{nocedal2006numerical}, it has traditionally been considered less
efficient than the standard full-space method because of increased nonzero
entries in the KKT system. When implemented on GPUs, however, it
offers the key advantage of ensuring positive definiteness in the
condensed KKT system. This, in turn, allows for the utilization
of linear solvers with a fixed numerical pivot sequence (known as
refactorization), an efficient implementation of which is available in the CUDA library.
Although this method is susceptible to numerical stability issues due to
the increased condition number in the KKT system, our results demonstrate
that the solver is robust enough to solve problems with a relative
accuracy of $\epsilon_{\text{mach}}^{1/4}\approx 10^{-4}$.

We present numerical benchmark results to showcase the efficiency of
our method, utilizing our two packages MadNLP.jl and
ExaModels.jl. The  KKT system is solved using the
external cuSOLVER library. To assess the performance of our method, we
compare it with standard CPU approaches using the data available in
pglib-opf \cite{babaeinejadsarookolaee2019power}.  Our benchmark results
demonstrate that our proposed computational framework has significant
potential for accelerating the solution of ACOPF problems, especially
when a moderate accuracy (e.g., $10^{-4}$) is sufficient.
Notably, when running on NVIDIA GPUs, our method achieves a
4x speedup compared with our solver running on CPUs for
the largest instance. Moreover, for the same instance, our approach
surpasses the performance of existing tools (such as Ipopt interfaced
with JuMP.jl) by an order of magnitude. This finding underscores the
importance of harnessing the power of GPUs for tackling
the computational challenges in power systems.

\paragraph*{Contributions}
We present, for the first time, a sparse nonlinear optimization
solution framework that can run entirely on GPUs, with all the
performance-critical data arrays residing exclusively on
GPUs. Additionally, we introduce the concept of SIMD abstraction for
NLP problems, which results in an efficient implementation of
GPU-accelerated parallel AD. Furthermore, we propose the condensed IPM
with an inequality relaxation strategy for the first time, enabling
the treatment of the KKT systems of NLPs without numerical pivoting,
thus allowing the solution of sparse, large-scale NLPs (with a
prominent example being ACOPFs) on GPUs.
